# R Functions and Workflow
## Workflow

### Scripts
So far, we've been using the command line interface in the console to type our programs. While this works, you might have noticed how annoying it can be to type longer programs in. Additionally, you're probably going to want to save your work at some point - and right now, you'd have to use Notepad or a similar program to save anything you've done.

Luckily, there's a better way. In the top left corner of RStudio, there's a menu button called "File". Click this, then click "New Project". If you click "New Directory", and then "New Project", you'll be able to create a folder where you can automatically store all of your R code and files. This will also create an R Project file, which you can load to return to where you left off the last time you closed RStudio.

Let's load the tidyverse again, now that we're in a new directory:
```{r}
library(tidyverse)
```

Now that you're working in a new directory, go back into "File" and hover over "New File". There's a lot of options, but right now we care about two of them: **R Scripts** and **R Notebooks**. Open one of each.

In your new script file, type the following:

```{r eval = FALSE}
ggplot(iris, aes(Petal.Length, Petal.Width)) + 
  geom_point()
```

Highlight everything and then press Cmd/Ctrl and Enter at the same time. A graph should appear in your Viewer window.

Whoops, looks like we forgot to color the points by species - add the color aesthetic to your plot.

```{r echo = FALSE}
ggplot(iris, aes(Petal.Length, Petal.Width)) + 
  geom_point(aes(color = Species))
```

It should already be clear what the advantage of working with R Scripts is - you can change pieces of your code quickly, without having to worry about retyping things into the console. You can also save and open your scripts (Cmd/Ctrl+S, Cmd/Ctrl+O), which makes working on big projects much easier.

Now change your code so it looks like this:
```{r}
a <- ggplot(iris, aes(Petal.Length, Petal.Width)) + 
  geom_point(aes(color=Species))
```

What we're doing here is _assigning_ the plot to ```a```. Now, anytime you call ```a```, the plot will appear - try it!

```{r}
a
```

Now add the following line under the first two:

```{r}
a + theme(legend.position = "top")
```

This will move the legend to the top of our graph, much as if we had included the ```theme()``` function in our original plot. Your program now has - for the first time in this course - two steps to it: the assignment step (where we make ```a``` a plot), and the print step (where we add our ```theme()``` and print the plot). While in an R script, there are three ways you can run the whole program:  

* Click the very top of the document and press Cmd/Ctrl+Enter once for each step
* Highlight the entire code and press Cmd/Ctrl+Enter to run it all at once
* While your cursor is anywhere in the script, press Cmd/Ctrl+Shift+Enter to run the whole program at once

That last method is usually the fastest and easiest. 

### Notebooks
While scripts are great, they do have some drawbacks. For instance, if you have to do more major and slower tasks - like loading datasets and libraries, or complicated math - you'll have to redo that step every time you want to run the whole program, which is a pain. Also, running a script pulls up the console window, which is a little bit of a headache.

For that reason, I tend to work in R Notebooks. Open your new notebook file, and you'll see a new welcome page! The welcome page has a lot of good information in it - you can delete everything after the second set of three dashes once you've read it.

Inside a notebook, you can make chunks by pressing Cmd/Ctrl+Alt+I. These chunks run as individual scripts, which you can run the exact same way by using combinations of Cmd/Ctrl, Shift, and Enter. Using notebooks can be a little more efficient than scripts, though, because it offers you the ability to split your code steps into multiple pieces, which can let you iterate on an idea faster than using scripts alone.

No matter which you prefer, you should aim to have one script or notebook per task you perform - don't just have one long, continous notebook for everything you're doing. Also, make sure you give everything a descriptive name - there's nothing worse than needing a file a month or so later and having to open every notebook you've ever made to find it!

It's also a good idea to make a new R Project, in a new folder, for each major project you start in on. These sorts of things might not matter too much to you while you're learning - but once you're doing more complicated things with R, having good habits like these are essential.

### Memory, Objects, and Names
Let's go back to when we assigned a plot to ```a```:

```{r}
a <- ggplot(iris, aes(Petal.Length, Petal.Width)) + 
  geom_point(aes(color = Species))
```

The ```<-``` symbol is the _assignment_ operator. We can use it to define the object ```a``` as all sorts of different objects:

```{r}
# Assign a the value 10
a <- 10
# Print out the object a
a

a <- c(1,50,200)
a

a <- "Hello, world!"
a

a <- geom_point(data = iris, aes(Petal.Length, Petal.Width, color = Species))
ggplot() +
  a
```

You'll notice that ```a``` is now listed in the upper-lefthand corner of RStudio, under the "Environment" tab. That's because a is now defined in memory - we can use it in any of our code, anywhere we want. In fact, you can even define ```a``` in one file and call it in another, so long as you've already run the code defining it in your current RStudio session. 

This is really cool for a lot of reasons - it lets us do more complicated things with R - but can also cause some problems. If you keep defining objects with names like ```a```, it's easy to forget which variable stands for what - and so you can wind up making mistakes when using those variables later on. For instance, we just overwrote `a` 3 times in that last example - imagine if we had important data stored in there!

In order to avoid that sort of confusion, you should use descriptive names when creating objects. You should also decide on a standard way you're going to format those object names - some people prefer ```snake_case_names```, others ```use.periods```, and I personally prefer what's known as ```CamelCase```. Different organizations and groups have different preferred styles (here's [Google's](https://google.github.io/styleguide/Rguide.xml)), but what's important right now is that you pick a style that makes sense to you. Be consistent using this style whenever you code - R won't understand you if you mess up your capitalization!

By the way - you might remember that I mentioned last unit that `=` could also be used as an assignment operator. That's true, but you should try to never do it - it makes your code much harder to understand - for instance, compare these two formats:

```{r}
a <- 10
a <- a + 1

b = 10
b = b+1

a == b
```

R's telling us that these two formats do exactly the same thing - define a variable as 10, and then overwrite the variable as the original value, plus one. But the top block of code makes a bit more sense than the lower one, which looks like we're trying to test if `b` is equal to `b + 1`, which is nonsense. This is just one of the reasons it's usually better to use `<-` for assignments.


### Dataframes and Transformations
Earlier in this course, we went over the different classes of vectors - character, numeric, and logical. If you're ever trying to find out what class a vector belongs to, you can call the ```class()``` function:

```{r}
SampleVector <- c(1,2,3)
class(SampleVector)
```

Note that we don't put object names (such as the name of our vector) in quotes. The general distinction is that if something exists in the global environment, we don't put it in quotes. If it isn't, we do. You can see what's in the current environment by looking at the "Environment" tab that I mentioned earlier - that tab is a list of all the objects you've defined so far in this session. Remember that even though your installed packages aren't in that list, you still don't put them in quotes when you call ```library()```.

A martix made of vectors is known, in R, as a ```dataframe```. We've already seen some simple dataframes in the past unit built using ```data.frame```:

```{r}
data.frame(x = c(1,2,3),
           y = c("a","b","c"),
           z = c(TRUE, TRUE, FALSE))
```

This is an example of something known as _rectangular data_ - the sort you're likely to find in spreadsheets and many, if not most, scientific applications. We'll be dealing with rectangular data almost exclusively in this course - while non-rectangular data is useful in many applications, it's much harder to get started with.

In fact, we'll almost always be working with a very specific type of rectangular data known as _tidy data_. Tidy dataframes always take the same shape:

```{r echo = FALSE}
data.frame("." = c("Observation 1", "Observation 2","...", "Observation n"),
           "Variable_1" = c("Value", "Value", "...", "Value"),
           "Variable_2" = c("Value", "Value", "...", "Value"),
           "Variable .." = c("Value", "Value", "...", "Value"),
           "Variable_n" = c("Value", "Value", "...", "Value"))
```

Tidy data is organized as follows:

* Each column is a single **variable**
* Each row is a single **observation**
* Each cell is a single **value**

As you might guess from the name, the `tidyverse` is specifically designed to work with tidy datasets. By storing all data in this format, we're able to quickly apply the same sets of tools to multiple different types of data. For instance, imagine a dataframe of seasonal temperatures, built as such:

```{r}
SeasonalTemps <- data.frame(Year = c(2015, 2016, 2017, 2018),
           Winter = c(40, 38, 42, 44),
           Spring = c(46, 40, 50, 48),
           Summer = c(70, 62, 81, 76),
           Fall = c(52, 46, 54, 56))
SeasonalTemps
```

This dataframe makes some sense - it's pretty easy to understand as a human reader, and would probably be a good layout for a printed table. But the problems with this format become obvious when we, for instance, try to graph the data:

```{r}
ggplot(SeasonalTemps, aes(x = Year)) + 
  geom_line(aes(y = Winter), color = "purple") + 
  geom_line(aes(y = Spring), color = "green") + 
  geom_line(aes(y = Summer), color = "blue") + 
  geom_line(aes(y = Fall), color = "red")
```
What a mess! That took far too long to type - a good general rule of thumb is that if you have to repeat yourself more than twice to do something, there's a better way to do it. And, even after all our effort, our graph doesn't have a legend, and the Y axis is labeled wrong.

Luckily enough, the ```tidyverse``` contains a package designed for making our data tidier - called, helpfully enough, ```tidyr```. We already loaded this package when we called the tidyverse earlier.

tidyr provides two essential functions for "reshaping" the data - changing back and forth between the _wide_ format we used above and a _long_ format, easier used by our functions. To change our ```SeasonalTemps``` data to a long format, we can use the ```gather()``` function. This function _gathers_ values stores in multiple columns into a single variable, and makes another variable - the _key_ variable - representing what column the data was originally in.

```gather()``` takes three important arguments:

* ```data```, the dataframe to gather  
* ```key```, what to name the key column   
* ```value```, what to name the column data was merged into

Additionally, we can specify columns that we want to preserve in the new, long dataframe by putting ```-ColumnName``` at the end of the function.

What this looks like for our seasonal data is something like this:

```{r}
LongTemps <- gather(data = SeasonalTemps, key = Season, value = AvgTemp, -Year)
LongTemps
```

Note that you don't have to type ```data = ```, ```key = ```, and ```value = ``` - if you don't, R assumes that you've listed the arguments in this order. 

This format makes graphing significantly easier:

```{r}
ggplot(LongTemps, aes(x = Year, y = AvgTemp, color = Season)) + 
  geom_line()
```

If, after all our hard work, we want to get back to our original wide format, we can undo our ```gather()``` using ```spread()```. Again, I'm giving spread a data, key, and value argument - but this time, the function is making a new column for each value of our key:

```{r}
WideTemps <- spread(LongTemps, Season, AvgTemp)
WideTemps
```

This new dataframe isn't quite the same as our original - the columns are now in alphabetical order! If we wanted to rearrage them, I find the easiest way is using the ```select()``` function from ```dplyr()```, another package in the tidyverse. By giving ```select()``` an argument for data and a vector of column names, we can rearrange the order the columns appear:

```{r}
OrderWideTemps <- select(WideTemps, c(Year, Winter, Spring, Summer, Fall))
OrderWideTemps
```

When doing this, though, we have to be careful we don't accidentally forget a column:

```{r}
select(WideTemps, c(Year, Winter, Spring, Fall))
```

Although, if we wanted to drop a column, we can do so by using a ```-``` sign:

```{r}
select(WideTemps, -Summer)
```

### The Pipe
At this point, we've created four dataframes - ```SeasonalTemps```, ```LongTemps```, ```WideTemps```, and ```OrderedWideTemps``` - which all contain the same data. When repeatedly making similar but different dataframes, it can be hard to keep track of which object has which data - and it can be hard to keep coming up with simple, descriptive names, too. One solution could be to keep overwriting the same object with the new data:

```{r}
a <- 10
a <- a*2
a <- sqrt(a)
```

But this breaks our rule - that if you have to repeat yourself more than twice, there's a better way to do it. Plus, if you make a mistake while writing over a value that had your original data in it, you have to start all over again - assuming that your data was saved anywhere else!

Luckily, the tidyverse also introduces a new operator ```%>%```, called the pipe. What the pipe does is pretty intuitive - it takes the output of whatever's on the _left_ side of the pipe, and uses it as the first input to whatever's on the _right_ side. For instance:

```{r}
Numbers <- c(5,10,15,20,25)

Numbers %>%
  mean()
```

Since all of the tidyverse functions take ```data``` as their first argument, this lets us _chain_ together multiple functions and skip those assignment steps:

```{r}
LongTemps %>%
  spread(Season, AvgTemp) %>%
  select(-Summer)
```

This makes our code much more easy to understand than constantly using the ```<-``` operator, plus it's an improved way to perform multiple steps in a way that's harder to make serious mistakes doing.

Even when a function doesn't have data as its first input, you can still use a pipe by typing ```data = .``` into the function:

```{r}
LongTemps %>%
  spread(data = ., Season, AvgTemp) %>%
  select(-Summer)
```

And pipes work well with ggplot2, too:

```{r}
LongTemps %>%
  ggplot(aes(x = Year, y = AvgTemp, color = Season)) + 
  geom_line()
```

### Data Transformations 

#### Mutate
This becomes useful when we want to transform our data itself for a graph, rather than transform the axes. For example, remember how we made our log-log graph last unit?

```{r}
LongTemps %>%
  ggplot(aes(x = Year, y = AvgTemp, color = Season)) + 
  geom_line() + 
  scale_y_log10()
```

This is useful, but ggplot only has a certain number of transformations built in (type ```?scale_y_continuous()``` for more info). Additionally, sometimes we'll want to transform our data for analyses - not just graphing. For this purpose, we can use ```dplyr```'s ```mutate()``` function. Mutate takes three arguments: the dataframe (which it can get from ```%>%```), the name of your new column, and what value the new column should have. Say, for example, we wanted to multiply our average temperatures by two:

```{r}
LongTemps %>%
  mutate(TwiceTemp = AvgTemp * 2)
```

You can make multiple columns in the same ```mutate()``` call:

```{r}
LongTemps %>%
  mutate(TwiceTemp = AvgTemp * 2,
         TwiceSquaredTemp = TwiceTemp^2,
         YearSeason = paste(Year, Season))
```

Notice I used a new function, ```paste()```, for that last column. This function pastes together values into a single cell - it can use other values in a dataframe, vectors, or strings. For instance:

```{r}
LongTemps %>%
  mutate(YearSeason = paste("The", Season, "of", Year))
```

Anyway. 

If you're transforming your data and don't want to save the old column, use ```transmute()```:

```{r}
LongTemps %>%
  mutate(TwiceTemp = AvgTemp * 2) %>%
  transmute(TwiceSquaredTemp = TwiceTemp^2,
         YearSeason = paste(Year, Season))
```

#### Tibbles
As I mentioned earlier, data in R is stored in _dataframes_. However, you may have noticed that the dataframe outputs from tidyverse functions look pretty different in your R session (I'd even say nicer) than our raw datasets! That's because of another useful tidyverse package, ```tibble```. 

Of course, the outputs in this book are pretty much the same - the technology I'm using to publish this isn't quite that advanced, yet.

We don't need to get too far into the mechanics of this package - if you load the tidyverse, any new dataframes you make will be converted into tibbles by default. If you want to force a dataframe into this format, use ```as.tibble()```; if you need the basic dataframe, use ```as.data.frame()```.

#### Subsetting Data

Let's go back to our ```iris``` dataset. I'm going to turn it into a tibble and then view it:

```{r}
iris <- as.tibble(iris)
iris
```

If we only wanted to work with part of this dataset, R gives us a lot of options to _subset_ the data. For instance, if we only wanted the first column containing sepal length, we could type this:

```{r}
iris[, 1]
```

If we wanted the first row, meanwhile, we'd type this:

```{r}
iris[1, ]
```

If we wanted several rows, we can specify them with ```c()``` or, if they're consecutive, ```:```. For instance:

```{r}
iris[c(1,2,3,4), ]
iris[1:4, ]
```


And if we wanted the value in the first row of the first column, we'd type this:

```{r}
iris[1,1]
```

The pattern should be clear now - inside of the braces, you type the row number, a comma, and then the column number. Notice that ```[]``` always gives us a tibble (or dataframe) back. If we wanted a vector, we could use ```[[]]```:

```{r}
iris[[1, 1]]
```

If we want to use column names instead of numbers, we could use ```$``` in the place of ```[[]]``` - note that this always returns a vector, not a dataframe:

```{r}
iris$Sepal.Length
```

The ```$``` is really helpful in using other base R functions:

```{r}
mean(iris$Sepal.Length)
sd(iris$Sepal.Length)
cor.test(iris$Sepal.Length, iris$Sepal.Width)
```

(Note that "cor.test()" runs Pearson's correlation test for whatever vectors you feed it - more on that test later, or [here](https://bookdown.org/ndphillips/YaRrr/correlation-cor-test.html)).

And ```$``` also lets us filter our data with conditionals - getting values that are equal to something, larger or smaller than it, and so on. For instance, if we want a dataframe (so ```[]```) where the rows (```[, ]```) all have a Species value of (```==```) "setosa":

```{r}
iris[iris$Species == "setosa", ]
```

Note that the species name is in quotes, because it's a character string. We don't have to do that for numeric values:

```{r}
iris[iris$Sepal.Length > 7.5, ]
```

You can use ```==```, ```>```, ```>=```, ```<```, ```<=```, and ```!=``` (not equal) to subset your data.

#### Filtering with the Tidyverse
This code is hard to read as a human, and doesn't work well with other functions. For instance, imagine trying to make a scatterplot of just the setosa data - your code will become almost unparseable. 

Instead, for more involved subsets, dplyr has a useful ```filter()``` function. It takes two arguments - your dataframe and the condition it should filter based on:

```{r}
iris %>%
  filter(Species == "setosa")
```

```filter()``` can use all the same operators as the ```[]``` methods of subsetting. Additionally, you can use ```&``` ("and") and ```|``` ("or") to chain filters together:

```{r}
iris %>%
  filter(Species == "setosa" & Sepal.Length == 5.1 & Sepal.Width == 3.3)
```

It's important to remember that ```&``` means things which satisfy EACH condition. A common mistake is to type:

```{r}
iris %>%
  filter(Species == "setosa" & Species == "versicolor")
```

Which, because no flower is both species, returns nothing.

In this case, you can either use an ```|``` ("or") operator, or - particularly if you have several cases you want to accept - ```%in%```:

```{r}
iris %>%
  filter(Species %in% c("setosa",
                        "versicolor"))
```

So long as your species is ```%in%``` the vector ```c()``` you provide, it will show up in the output.

### Working with Groups

Say we wanted to find the mean sepal length in our dataset. That's pretty easy:

```{r}
mean(iris$Sepal.Length)
```

But we already know from our graphs that sepal length differs dramatically between species. If we wanted to find the mean for each species, we could calculate it individually for each group:

```{r collapse = TRUE}
setosa <- iris %>%
  filter(Species == "setosa")
virginica <- iris %>%
  filter(Species == "virginica")
versicolor <- iris %>%
  filter(Species == "versicolor")


mean(setosa$Sepal.Length)
mean(virginica$Sepal.Length)
mean(versicolor$Sepal.Length)
```

But that code is messy, the output is without any context, and it goes against our rule - that if you have to repeat yourself more than twice, there's a better way to do it.

The better way in the tidyverse is to use _grouping_ and _summary_ functions. In the following example, we'll use ```group_by()``` to group our dataframes by the species types, and ```summarise()``` to calculate the mean for each of them (in a column called "MeanSepalLength"):

```{r}
iris %>%
  group_by(Species) %>%
  summarise(MeanSepalLength = mean(Sepal.Length))
```

This is a faster and easier to understand way to perform functions on groups of data. Note that ```summarise()``` uses the British spelling - almost all functions in R have British and American spellings built in (you can use ```color``` or ```colour``` aesthetics in ggplot, for instance), but this is an important exception. While there is a function called ```summarize()```, it's highly glitchy and its use is highly discouraged.

You can use ```group_by()``` to calculate all sorts of things - for instance, we can calculate the distance of each plant's sepal length from the group mean, as follows:

```{r}
iris %>%
  group_by(Species) %>%
  mutate(SLDistanceFromMean = Sepal.Length - mean(Sepal.Length))
```

If you want to calculate variables for the whole dataset again, you'll have to ungroup your data - dataframes will stay grouped until you actively ungroup them with ```ungroup()```. For instance, to calculate the distance of each plant's sepal length from the overall mean:

```{r}
iris %>%
  select(c(Sepal.Length, Species)) %>%
  group_by(Species) %>%
  mutate(SLDistanceFromGroupMean = Sepal.Length - mean(Sepal.Length)) %>%
  ungroup() %>%
  mutate(SLDistanceFromTotalMean = Sepal.Length - mean(Sepal.Length))
```

(Note that I got rid of some columns with ```select()``` to make all the columns in the tibble fit on one page.)

### Missing Values

#### Explicit Missing Values
Working with data, there are often two types of missing values we have to worry about. The obvious one are _explicit_ missing values, represented in R as ```NA``` (or, sometimes, ```NaN```). Let's make a dataframe:

```{r}
MissingExample <- tibble(w = c(1, 2, 3),
                         x = c("A", "B", "C"),
                         y = c("do", "re", NA),
                         z = c(807, NA, 780))
MissingExample
```

(I'm using ```tibble()``` in place of ```dataframe()``` here, but the outcome is almost identical.)

```NA``` values are a little tricky to work with - look what happens when we try to find the mean of ```z```:

```{r}
mean(MissingExample$z)
```

The reason this happens is because we _don't know_ what the mean is - that NA value could be anything, so it's impossible to know what the mean is. To get around this, we can set the ```na.rm``` argument to ```TRUE```:

```{r}
mean(MissingExample$z, na.rm = TRUE)
```

We can also solve the problem with filtering out the ```NA``` values. We can use ```is.na()``` to find out where certain values are, and then ask ```filter()``` to remove those rows from our dataset as follows:

```{r}
MissingExample %>%
  filter(!is.na(z)) %>%
  summarise(Mean = mean(z))
```

```!``` means "negation" in R, or "opposite" - so we're asking ```filter()``` to return the _opposite_ of any row where ```z``` is ```NA```, or, alternatively, all the rows where it has a value.

If we wanted to drop _every_ row that has a ```NA```, we could use the following ```tidyr``` function:

```{r}
MissingExample %>%
  drop_na()
```

Or, if we knew the values we wanted those ```NA``` to represent, we could use ```replace_na()```, also from ```tidyr```. We just have to specify a list of what we want those values to be:

```{r}
MissingExample %>%
  replace_na(list(y = "mi", z = "078"))
```

Notice a difference in the ```z``` column with this example? Because I put "078" in quotes, it changed the entire column to a character vector - because quotes mean characters, and a vector can only hold one class of data.

We'll talk more about that ```list()``` function later on - that's a little too complicated for this unit.

#### Implicit Missing Values
The other, harder to identify type of missing value is the _implicit_ missing value. Say we have a dataframe ```TreeData```, which lists the species that are present at two different sites:

```{r}
TreeData <- tibble(Site = c("A","A","A","B","B"),
                   Species = c("Red Maple", "Sugar Maple", "Black Cherry", "Red Maple", "Sugar Maple"),
                   Count = c(10,5,15,8,19))
TreeData
```

A lot of field data is collected like this, where each row represents something that was present at the field site. The problem with this comes when we try to calculate summary statistics for each species:

```{r}
TreeData %>%
  group_by(Species) %>%
  summarise(Mean = mean(Count), StandardDev = sd(Count))
```

Black cherry has a missing (```NaN```) standard deviation, because as far as R knows, it only has one observation to make estimates with. In reality, the fact that black cherry was missing from site B is a data point in and of itself - it's an _implicit_ value of 0.

To fix that, we can use the ```complete()``` command from ```tidyr```. This function takes column names as arguments, and returns a dataframe with every combination of the values in those columns. We can also specify what to replace ```NA``` values with, much like we did in ```replace_na()```, with ```fill```:

```{r}
TreeData %>%
  complete(Site, Species, fill = list(Count = 0))
```

This way, when we go to calculate our summary statistics, we get better answers:

```{r}
TreeData %>%
  complete(Site, Species, fill = list(Count = 0)) %>%
  group_by(Species) %>%
  summarise(Mean = mean(Count), StandardDev = sd(Count))
```

### Count Data
One other common issue with field data is that it's in a summary form - for instance, our tree data summarizes the number of trees at each site into one column. This is often easier to record in the field and easier to read as a human - but it makes some analyses much harder!

The function ```uncount()``` makes this pretty easy for us:

```{r}
LongTreeData <- TreeData %>%
  uncount(Count)

LongTreeData
```

And if we wanted to get back to the summary table, we can use ```count()```:

```{r}
LongTreeData %>%
  count(Site, Species)
```

If we want to change that column ```n```'s name to something more descriptive, we can use ```rename()```:

```{r}
LongTreeData %>%
  count(Site, Species) %>%
  rename(Count = n)
```


### Oddballs

Predict what will happen when you run the following code - then run it!

```{r eval = FALSE}
sqrt(2)^2 == 2

1/49 * 49 == 1
```

You'd expect both of these things to be true, but R seems to think otherwise.

That's because R has to estimate the true value of things like 1/49 - it only calculates to so many digits, because it can't store an infinite number of decimal places. As such, 1/49 * 49 _isn't_ exactly equal to 1 - it's just near it. To catch these sorts of things, use ```near()``` instead of ```==```:

```{r, collapse = TRUE}
1/49 * 49 == 1
near(1/49 * 49, 1)
```

## R Functions and Workflow Exercises

### Do the following:
1. What class is the vector ```c(1, TRUE, 3)```? Why is it not a character vector?
2. Make and print this tibble. What do the abbreviations under each column name mean?
```{r eval = FALSE}
tibble(x = c(1, 2, 3),
       y = c("A", "B", "C"),
       z = c(TRUE, FALSE, TRUE))
```
3. Inspect the ```smiths``` dataset (loaded with the tidyverse - you can access it like ```iris```). How can you fix those missing values?

### Work with other datasets:
1. ```spread()``` the ```iris``` dataset so that each species' petal width is in its own column. Then ```gather()``` the table back together. What's different about this dataframe?
2. Select all the rows of ```iris``` where the species is setosa. Now select all the rows where the species _isn't_ setosa.
3. What's the mean price for each cut of diamond in the ```diamonds``` dataset? 
 