# Unit 2: R Workflows
## This is a work in progress - only parts 1 and 2 are ready for other people to look at
So far, we've been using the command line interface in the console to type our programs. While this works, you might have noticed how annoying it can be to type longer programs in. Additionally, you're probably going to want to save your work at some point - and right now, you'd have to use Notepad or a similar program to save anything you've done.

Luckily, there's a better way. In the top left corner of RStudio, there's a menu button called "File". Click this, then click "New Project". If you click "New Directory", and then "New Project", you'll be able to create a folder where you can automatically store all of your R code and files. This will also create an R Project file, which you can load to return to where you left off the last time you closed RStudio.

Let's load the tidyverse again, now that we're in a new directory:
```{r}
library(tidyverse)
```

Now that you're working in a new directory, go back into "File" and hover over "New File". There's a lot of options, but right now we care about two of them: **R Scripts** and **R Notebooks**. Open one of each.

In your new script file, type the following:

```{r eval = FALSE}
ggplot(iris, aes(Petal.Length, Petal.Width)) + 
  geom_point()
```

Highlight everything and then press Cmd/Ctrl and Enter at the same time. A graph should appear in your Viewer window.

Whoops, looks like we forgot to color the points by species - add the color aesthetic to your plot.

```{r echo = FALSE}
ggplot(iris, aes(Petal.Length, Petal.Width)) + 
  geom_point(aes(color=Species))
```

It should already be clear what the advantage of working with R Scripts is - you can change pieces of your code quickly, without having to worry about retyping things into the console. You can also save and open your scripts (Cmd/Ctrl+S, Cmd/Ctrl+O), which makes working on big projects much easier.

Now change your code so it looks like this:
```{r}
a <- ggplot(iris, aes(Petal.Length, Petal.Width)) + 
  geom_point(aes(color=Species))
```

What we're doing here is _assigning_ the plot to ```a```. Now, anytime you call ```a```, the plot will appear - try it!

```{r}
a
```

Now add the following line under the first two:

```{r}
a + theme(legend.position = "top")
```

This will move the legend to the top of our graph, much as if we had included the ```theme()``` function in our original plot. Your program now has - for the first time in this course - two steps to it: the assignment step (where we make ```a``` a plot), and the print step (where we add our ```theme()``` and print the plot). While in an R script, there are three ways you can run the whole program:  
* Click the very top of the document and press Cmd/Ctrl+Enter once for each step
* Highlight the entire code and press Cmd/Ctrl+Enter to run it all at once
* While your cursor is anywhere in the script, press Cmd/Ctrl+Shift+Enter to run the whole program at once

That last method is usually the fastest and easiest. 

While scripts are great, they do have some drawbacks. For instance, if you have to do more major and slower tasks - like loading datasets and libraries, or complicated math - you'll have to redo that step every time you want to run the whole program, which is a pain. Also, running a script pulls up the console window, which is a little bit of a headache.

For that reason, I tend to work in R Notebooks. Open your new notebook file, and you'll see a new welcome page! The welcome page has a lot of good information in it - you can delete everything after the second set of three dashes once you've read it.

Inside a notebook, you can make chunks 