# More Complicated Analyses
## Work in Progress
## Other Datasets
### Importing Your Own Data
So far in this course, we've been working exclusively with the data pre-available in R and a few of the packages we've taken advantage of. While this is a good way to learn R, it's not particularly helpful in letting you complete your own analyses.

Luckily, R has a number of functions used to import data from external files. To demonstrate these, I'll be using datasets located [on GitHub](https://github.com/mikemahoney218/Unit6Data), in the "Datasets" folder. You don't necessarily need to download these files - but we'll be using these for demonstration throughout this unit.

We're going to be assuming that your data are located in the same folder as your script - this is usually the easiest way to manage datasets. Otherwise, you'll have to wrestle a little with using relative pathways and directory names - Hadley Wickham explains these in more detail here.

In order to do this, we'll be using the ```readr``` package, which is included in the base tidyverse:

```{r}
library(tidyverse)
```

All the datasets included in our example use different _delimiters_ - the character that tells R (or Excel, or whatever program you're using) where one column stops and the next one begins. No matter what your delimiter is, we can parse the file using ```read_delim()``` - the first argument is the filename to read, while the second is the delimiter used. For instance, our text file is tab delimited - and since tabs are represented in R as ```\t```, we'll tell ```read_delim()``` to use that as a delimiter:

```{r}
SiteData <- read_delim("SiteData.txt", "\t")
SiteData
```

We can do the same thing with documents that have comma separated values (known as CSVs):

```{r}
CoverData <- read_delim("CoverData.csv", ",")
CoverData
```

However, ```readr``` also includes a pretty good tool specifically for CSV files:

```{r}
CoverData <- read_csv("CoverData.csv")
CoverData
```

Now, ```readr``` doesn't have native support for more complicated files, like Excel files. Philosphically, you shouldn't store data in Excel format for long periods of time - we have no idea how long Microsoft will be around for, and the encoding used by Excel may someday disappear and take your data with it. CSVs are generally preferred for long-term data storage, as they're easy to understand visually and are easily parsed by computers.

However, data entry is much easier in Excel, and plenty of data professionals still use the format. Luckily, there's a package - ```readxl``` - designed to parse these types of files:
```
install.packages("readxl")
```

```{r}
library(readxl)
```

```{r}
PlotData <- read_excel("PlotData.xlsx")
PlotData
```

You'll also notice that RStudio has an "import dataset" button in the top right corner, which makes use of both the ```readr``` and ```readxl``` packages. This button lets you point and click your way through data import, and then copy the code into your script. It's a great resource for beginners!

### Exporting Data
Writing data to a file is also pretty painless using ```readr```. There are as many options for delimiters as before - you can use ```write_delim()``` to specify which you want to use with your data - but more commonly data is imported and exported as CSV files using ```write_csv()```

```{r}
write_csv(CoverData, "CoverData.csv")
```

Note that, like all other ```tidyverse``` functions, the dataset is the first thing you specify.


INCLUDE JOINS
AND AIC


```{r}

```

