# Playing Nicely With Others
## Work in Progress

So far, we've gotten ourselves to a point where we can explore our data, perform some analyses, and graph our results intelligently. That's all great, and are incredibly important skills. However, even more important than the hard coding skills we've been building is the ability to communicate our results - to our peers, bosses, and the larger outside world.

We'll be using a lot of non-R technologies this unit, and focusing mostly on how to use these technologies in a workflow that resembles those used in the modern scientific world. Plenty of scientists don't use R Markdown. Even more don't use Git. Some may have never heard of LaTeX. But these are skills and tools which are growing in importance and acceptance throughout the scientific and business worlds, so getting used to them now will make your life significantly better. But even moreso than that, understanding how to learn tools like these - at least to the depths you need them - will help you pick up new tools and tech as the world continues to advance. We aren't going super deep into any tool - but we'll go deep enough to make their benefits obvious.

## R Markdown
The Markdown programming language was developed in 2004 to let developers write HTML in a way that is human-readable and easily writeable. R Markdown is an implementation of that language in R, which lets you embed R code and graphs seamlessly into your documents.

To get more information on R Markdown, check out Yihui Xie's [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/), written by the guy who made the format.

We can start a new Markdown document in the same place we make new scripts and notebooks - click File -> New File -> R Markdown. This will open up a window with some default templates - Markdown can be used to make web apps, presentations, and even entire websites. You'll also see that a number of journal formats are pre-loaded in the templates option. For now, let's just open a blank document, and have it output as an HTML file.

You'll notice a lot of pre-filled in the document. At the very top (inside two lines of ```---```) is what's known as the [YAML](https://en.wikipedia.org/wiki/YAML) header. This is where you can set a number of top-level options for your document, including font and margin sizes. This is also where you should specify your document title, author name, and date (using the ISO format: YYYY-MM-DD).  

We won't mess too much with the header right now, but I do want to point out one important difference between YAML and the R code we've gotten used to. In R, we use parentheses to specify which arguments belong to which function. In YAML, we use spacing, with an indentation of two spaces taking the place of parentheses. For instance, if we wanted to add a table of contents to our HTML document, we could specify that with ```toc = TRUE``` in R as follows:

```
html_document(toc = TRUE)
```

In YAML, those parentheses get replaced by spacing:

```
output:
  html_document:
    toc: true
```

So the number of spaces you use matter in YAML, while they don't in R. This still trips me up, which is why I emphasize it now!

Now, you might notice that the Markdown document looks a lot like the R Notebooks we've been experimenting with throughout the course. That's because they're actually the same thing - notebooks are a specialized implementation of Markdown, which let you run chunks of code faster and easier than a markdown document. As such, you can write an entire document in a notebook, and then move it to Markdown once you're looking to share the document around. All of the options we'll cover below still apply to notebook documents, and you can still make chunks of code in Markdown by pressing Ctrl+Alt+I.

The Markdown format lets you stylize your text in a way that doesn't impede readability and won't require reformatting every time page numbers change, for instance. It does take a second to learn, but luckily there are a number of [cheatsheets and reference guides](https://www.rstudio.com/resources/cheatsheets/) available as resources. 

There's a ton of tools available for stylizing your documents. Here, we're going to quickly demonstrate the most commonly used ones:

* `#` with a space after it makes headers - the more `#` you have in a row, the smaller the header.
* `*` with a space after it makes a bulleted list
* Numbers with a `.` or `)` after them start a numbered list. You can also use `#)` or `#.` if you don't want to worry about numbering the list yourself.
    * For both types of list, you need an empty line of whitespace before the first bullet to make a list.
    * By the way, putting four spaces before the `*` lets you make an indented bullet.
* Two spaces at the end of a line of text  
start a new paragraph. 
If you only use one space, both lines will print continuously on a single line.
* You can put spaces between paragraphs by entering a single line of 

white space.  

* One set of `_` or `*` on both sides of a word makes it *italicized*
* Two sets of `__` or `**` on both sides of a word make it **bold**
* Superscripts can be made using `^` around te^x^t
* Subscripts can be made using `~` around te~x~t
* Links are made following the format `[text](link)`
* You can make block quotes using `>` before the text

You can also use <font color = "red">HTML</font> directly in the body of your Markdown documents, if you need more specialized formats.

When you want to view the output of your document, press the `Knit` button on the toolbar to turn your Markdown file into whatever output file type you've specified. You can then share the resulting file - in our case, an HTML document - with whatever stakeholder needs to see it.

There's plenty more depth to be discovered in how to best format your R Markdown documents - for instance, there are extensions to automatically add bibliographies and figure captions to your entire document - but we'll leave those for another reader to cover. For now, the point in us going over how to use R Markdown is to enable you to communicate your code better, by making documents that explain what each of your code chunks is doing - and, more importantly, why. By having documents that clearly label and show each of the pieces of your code next to their outputs, researchers and programmers are better able to communicate their results and methodologies with one another.

### Kable
Now that we know how to generate pretty figures and text, it's time we learn how to make attractive tables, too. We'll be using the ```iris``` dataset as an example of how to use the `kableExtra` package to do just that. Note that most of this information comes from the [kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html) documentation, which I've found very easy to use.

First off, let's load the tidyverse:
```{r}
library(tidyverse)
```


When you print out tables in a Markdown document, the generic output looks a lot like the plain text I've been using in the past units:
```{r}
head(iris)
```

However, there happens to be a better way. Included in the `knitr` package is the function `kable()`, which makes attractive-ish tables:

```
install.packages("knitr")
```
```{r}
library(knitr)
```

```{r}
head(iris) %>%
  kable()
```

In my mind, though, a more attractive option is provided by the `kable_styling()` function from the `kableExtra` package:

```
install.packages("kableExtra")
```
```{r}
library(kableExtra)
head(iris) %>%
  kable() %>%
  kable_styling()
```

Once we have that table to work from, we're able to customize it with a number of different options. For instance, if we want our table to have striped rows, which get darker when you hover a cursor over them, are less wide (more condensed) than usual, and are responsive to how big the user's screen is, and we don't want our table to take up the full screen width allowed, we can specify all those as options:
```{r}
head(iris) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", 
                                      "hover", 
                                      "condensed", 
                                      "responsive"), 
                full_width = FALSE)
```

Heck, if we wanted to show all our data, we can include a scroll bar, to keep the printout from being overwhelming:

```{r}
iris %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", 
                                      "hover", 
                                      "condensed", 
                                      "responsive"), 
                full_width = FALSE) %>%
  scroll_box(height = "250px")
```

I'm not walking through all the options as in-depth as I usually will, because I think the [documentation](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html) for this package is extremely well-done. I just wanted to give you the skills to control what your tables look like as much as you may want.

## LaTex

LaTeX (pronounced Lay-Tek) is another document creation language, much like Markdown. We won't go too far into the minutae of LaTex, as it can be a little intimidating. We'll only be looking at the parts of LaTex that we can use directly in Markdown, simply by putting dollar signs (`$`) around the LaTex command. 

Most commonly, these commands are used to make attractive equations or include special symbols, such as Greek letters and mathematical operators. Those operators are drawn by putting a `\` before the name of the symbol - for instance, the formula `$$\pi * 2^4 * \sqrt{27}$$` would draw the equation:

$$\pi * 2^4 * \sqrt{27}$$

For the capitalized version of Greek letters, just capitalize the first letter of the name - `$\delta$` generates $\delta$, versus `$\Delta$` generating $\Delta$, for instance.

Two dollar signs will start the equation (or symbol) centered on the next line, while one will generate the equation or symbol continuing on the same line.

That's about all the Latex we'll have a need to go over in this course. If you're interested in learning more, there's a pretty good tutorial available on [OverLeaf](https://www.overleaf.com/learn/latex/Learn_LaTeX_in_30_minutes), and a publicly available [cheatsheet](https://www.rstudio.com/resources/cheatsheets/).

## Git(Hub)
Git is a version control system originally developed in 2005 to let a lot of extremely talented coders work together on one of the largest, most complex open source projects the world has ever seen. It is a nightmare to learn, and every company does it slightly differently.

GitHub is a company founded in 2008, designed to provide streamlined servers and plugins to allow teams and non-professionals to use Git on various projects. The tools it provides are less powerful than Git itself, but much faster to learn.

For that reason, we'll be using GitHub - and the tools developed specifically for using Git with GitHub - exclusively in this unit. There's really no reason to learn Git by itself right now; it's a good skill to pick up eventually, but not while busy learning another computer language. If you're interested in a tutorial written specifically for Git, check out [this blog](http://kbroman.org/github_tutorial/) by Karl Broman.



## Commenting Code

https://uc3.cdlib.org/2014/05/05/github-a-primer-for-researchers/